% !TEX root = msri731.tex

\section{\texorpdfstring{$p$}{p}-adic Hodge theory for rigid spaces 1}
\thanksauthor{Wieslawa Niziol (Feb.\ 20)}





We will be discussing the finiteness of \'etale cohomology modulo $p$, and some 
basic comparison theorems. 





\subsection{Comparison theorem}

The main theorem is the following:

\begin{theorem}\label{thm:main-comp-thm}
Let $C$ be an algebraically closed complete extension of $\dQ_p$. Let $X$ be 
a proper smooth adic space over $\spa(C,\cO_C)$. Let $\sL$ be an $\dF_p$-local 
system on $X$. Then 
\begin{enumerate}
  \item $\h^i(X_\et,\sL)$ is a finite-dimensional $\dF_p$-vector space for all $i\geqslant 0$ 
  \item there is an almost isomorphism (of $\cO_C$-modules) 
    \[
      \h^\bullet(X_\et,\sL)\otimes_{\dF_p} \cO_C/p \simeq \h^\bullet(X_\et,\sL\otimes \cO_X^+/p) .
    \]
\end{enumerate}
\end{theorem}

There is a relative version of this theorem, which follows from the base change theorem 
and a slightly more general version of the main theorem. Moreover, this theorem implies 
the $\bdr$-comparison theorem. 

The proof works by redution to coherent cohomology via Artin-Schreier sequences. This 
involves tilting and the pro-\'etale site. Recall that the pro-\'etale site $X_\proet$ 
computes \'etale cohomology via $\nu:X_\proet \to X_\et$. Here we are mapping 
$\sF$ on $X_\et$ to $\mathsf R \nu_\ast \nu^\ast \sF$. 
Over $(K,K^+)$, affinoid perfectoids form a basis for the pro-\'etale topology. 
Finally, if $U$ is an affinoid perfectoid, then $\h^i(U,\sO_X^+/p)$ is almost 
zero for $i>0$. One does this by relating this cohomology group to 
$\h^i(W_\et,\sO_X^+/p)$ for $W$ an affinoid perfectoid space, and applying the 
almost version of Tate's acyclicity theorem. 

We will use a twisted version of the above theorem.

\begin{theorem}
Let $\sL$ be an \'etale $\dF_p$-local system. Then $\h^i(U,\sL\otimes \sO_X^+/p)$ is almost 
zero for $i>0$. Moreover, $\h^0(U,\sL\otimes \sO_X^+/p) = M(U)$, for 
$\widehat U=\spa(S,S^+)$, is a finitely generated projective $S^+/p$-module compatible 
with base change. 
\end{theorem}

We want to use the classical Artin-Schreier sequence on $X_\proet$. We have an 
exact sequence:
\[\xymatrix{
  0 \ar[r] 
    & \sL \ar[r] 
    & \sL\otimes \sO_X^+/p \ar[r]^-{\varphi-1} 
    & \sL\otimes \sO_X^+/p \ar[r] 
    & 0 
}\]
To see exactness, look locally at some affinoid perfectoid $U\in X_\proet$. We 
can assume $\sL_U$ is trivialized, so it is enough to show surjectivity. To see 
that, we use tilting. Let $\pi\in \sO_{\widehat U^\flat}^+$ be such that 
$\pi^\sharp=p$. 

(If $U=(U_i=\spa(A_i,A_i^+))\in X_\proet$, then $\widehat U$ comes from the 
completion of $\varinjlim A_i$.)


The map $\varphi-1:\sO_U^+/p \to \sO_U^+/p$ is \'etale, and is the same as 
$\sO_{U^\flat}^+/\pi \to \sO_{U^\flat}^+/\pi$. Pass back from the tilted site using 
the equivalence $\widehat U_\et^\flat \simeq \widehat U_\et$. We get 
\[\xymatrix{
  \cdots \ar[r] 
    & \h^i(X_\et,\sL) \ar[r] \ar@{.>}[d] 
    & \h^i(X_\et,\sL\otimes \sO_X^+/p) \ar[r] \ar[d]^-\wr 
    & \h^i(X_\et,\sL\otimes \sO_X^+/p) \ar[r] \ar[d]^-\wr 
    & \cdots \\
  & \dF_p^r \ar[r] 
  & (\cO_C/p)^r \ar[r] 
  & (\cO_C/p)^r \ar[r] 
  & 0
}\]
The isomorphism uses the fact (which we will prove later) that 
$\h^i(X_\et,\sL\otimes \sO_X^+/p)$ is an almost finitely generated $\cO_C/p$-module, 
as well as some facts about Frobenius. The vertical isomorphisms are only almost 
isomorphisms, and if we knew the appropriate finiteness conditions, we could 
pass from almost mathematics to ``real'' mathematics. 

Since we are trying to prove a finiteness theorem, we instead use a modified form 
of Artin-Schreier theory to get rid of the ``almost'' part. We have an exact sequence 
\[\xymatrix{
  0 \ar[r] 
    & \sL \ar[r] 
    & \sL\otimes \widehat \sO_{X^\flat} \ar[r]^-{\varphi-1} 
    & \sL\otimes\widehat\sO_{X^\flat} \ar[r] 
    & 0
}\]
where $\widehat\sO_{X^\flat}^+ = \varprojlim_\varphi \sO_X^+/p$ and 
$\widehat\sO_{X^\flat} = \widehat\sO_{X^\flat}\otimes_{\cO_{C^\flat}} C^\flat$. 
Passing to cohomology, we get 
\[\xymatrix{
  \cdots\ar[r] 
    & \h^i(X_\proet,\sL) \ar[r] \ar[d]^-\wr 
    & \h^i(X_\proet,\sL\otimes \widehat\sO_{X^\flat}) \ar[r] \ar[d]^-\wr 
    & \h^i(X_\proet,\sL\otimes \widehat\cO_{X^\flat}) \ar[r] \ar[d]^-\wr 
    & \cdots \\
  0 \ar[r] 
    & \dF_p^r \ar[r] 
    & (C^\flat)^r \ar[r] 
    & (C^\flat)^r \ar[r] 
    & 0
}\]
The proof of Theorem \ref{thm:main-comp-thm} is nearly complete. It remains to 
show that $\h^j(X_\et,\sL\otimes \sO_X^+/p)=0$ for 
$j>2\dim X$. Put $d=2\dim X$. Look at the projection 
$\varepsilon:X_\et \to X_\text{an}$. Since the dimension of $X_\text{an}$ is also $d$, it 
suffices to show that $\mathsf R^j \varepsilon_\ast(\sL\otimes \sO_X^+/p)$ are almost 
zero for $j>d$. We will assume $X$ is affinoid, smooth, and has a ``good'' coordinate 
system $X\to T^d$, where $T^d$ is a $d$-dimensional torus, and that this map is the 
composition of a rational endomorphism with a finite \'etale map. 

Write $T^d=\spa(C\langle T_i^{\pm 1}\rangle,\cO_C\langle T_i^{\pm 1}\rangle)$. 
Let $\widetilde T/T$ be the affinoid perfectoid, 
$\widetilde T=\varprojlim_m T_m$, where 
$T_m=\spa(C\langle T_i^{\pm 1/p^m}\rangle,\cO_C\langle T_i^{\pm 1/p^m}\rangle)$. Let $\widetilde X=X\times_T \widetilde T$. This is a pro-covering of $X$. (Somewhere 
in here, we put $T^d=T$.) 
Then 
\[
  \mathsf R\Gamma(X_\proet,\sL\otimes \sO_X^+/p) = \mathscr C(\widetilde X/X,\mathsf R\Gamma(\widetilde X_\proet^\bullet,\sL\otimes \sO_X^+/p)) 
\]
But 
\begin{align*}
  \h^j(\widetilde X_\proet^{i/X},\sL\otimes \sO_X^+/p) 
    &= \h^j(\widetilde X\times \dZ_p^{d(i-1)},\sL\otimes \sO_X^+/p) \\
    &= \hom_\text{cont}(\dZ_p^{d(i-1)},\h^j(\widetilde X,\sL\otimes \sO_X^+/p))
\end{align*}
The $\h^j(\widetilde X,\sL\otimes \sO_X^+/p)$ are almost zero for $j>0$. It follows that 
$\h^j(X_\proet,\sL\otimes \sO_X^+/p)$ is almost isomorphic to 
$\h_\text{cont}^j(\dZ_p^d,M)$, where $M$ is some finitely generated projective 
$S^+$-module. 

The group $\dZ_p^d$ has cohomological dimension $d$, so everything vanishes (in the 
almost sense) if $j>d$. 





\subsection{Proof of auxiliary result}

Let's prove the following claim. Let $X$ be a proper smooth adic space over 
$\spa(K,\cO_K)$, where $K$ contains all $p$-power roots of unity and has 
characteristic zero. Then $\h^i(X_\et,\sL\otimes \cO_X^+/p)$ is a finitely generated 
$\cO_K$-module. The idea is to mimic the proof of finite-generation of coherent 
cohomology for rigid analytic spaces. Write $X=\bigcup V_i$, where the $V_i$ are 
affinoids with good coordinates. We also want a covering $X=\bigcup V_i'$, where 
each $V_i'\subsetneq V_i$. This gives us two spectral sequences 
\begin{align*}
  E_1^{m_1 m_2} 
    &= \bigoplus_{|j|=m_1+1} \h^{m_2}(V_{i,\et},\sL\otimes \sO_X^+/p) \Rightarrow \h^{m_1+m_2}(X,\sL\otimes \sO_X^+/p)  \\
  E_2^{m_1 m_2}(V) 
    &= \bigoplus \h^i(X,V_j',\sL\otimes \sO_X^+/p) \Rightarrow \h^{m_1+m_2}(X,-)
\end{align*}
It suffices to show that the image of the restriction map is a finitely generated 
$\sO_X$-module. 

\begin{lemma}
Let $V$ be a ``good'' affinoid, $V'\subsetneq V$ a rational subset. Then the image of 
$\alpha=\text{res}:\h^i(V_\et,\sL\otimes \sO_X^+/p) \to \h^i(V_\et',\sL\otimes \sO_X^+/p)$ 
is an almost finitely generated $\cO_K$-module. 
\end{lemma}
\begin{proof}
Write $V'\hookrightarrow V\to T$, where $T$ is a torus. Let $\widetilde T$ be the 
``perfectoidification of $T$.'' This induces covers $\widetilde V\to V$ and 
$\widetilde V' \to V'$. Let $\widehat{\widetilde V}'=\spa(S',{S'}^+)$ and 
$\widehat{\widetilde V}=\spa(S,S^+)$ be the associated adic spaces coming from 
completion. (Formally, $\widetilde T=\varinjlim spa(T^{1/p^m})$ and 
$\widehat{\widetilde T}$ is the completion. 

First we pass to the group coordinates, giving 
\[
  \alpha:\h_\text{cont}^i(\dZ_p^d,M) \to \h_\text{cont}^i(\dZ_p^d,M\otimes_{S^+/p} {S'}^+/p) .
\]
We can assume $M=S^+/p$. Next, we consider $(S_m,S_m^+)$, 
$(S_m',{S_m'}^+)$, which are the ``levels'' in the pro-\'etale cover 
$\widetilde T\to T$. The image of 
\[
  \h_\text{cont}^i(\dZ_p^d,(S_m^+\otimes_{R_m^+}R^+)/p) \to \h_\text{cont}^i(\dZ_p^d,S_m'\otimes \cdots)
\]
is finitely generated over $\cO_K$. 

The rest of the proof uses Hochschild-Serre to reduce to a map 
$S_m^+\otimes \h^i \to {S_m'}^+\otimes \h^i$. 
\end{proof}


